---
output: github_document
---

# Logistic Regression(logreg)
  

This package implements logistic regression using numerical optimization. The functions in this package allow for you to:obtain the least squares estimators. This is used as a starting point for optimization, bootstrap confidence intervals for each predictor in the logistic regression, plot the fitted logistic curve to the responses, obtain a confusion matrix on the logistic regression and print metrics calculated from the matrix, and have the ability to plot the given metrics from 0.1 to 0.9. 

First, install the package.
```{r}
devtools::install_github("R-4-Data-Science/Final_Project_Group7/logreg")
library(logreg)
```
    
In order to show examples, we will load data from class and format it for use.

```{r}
mydata <- read.csv("expenses.csv")

mydata$sex <- factor(mydata$sex, 
                     levels = c("male","female"), labels = c(0,1))
mydata$smoker <- factor(mydata$smoker, levels = c("yes","no"), labels = c(1,0))
mydata$region_northeast <- ifelse(mydata$region=="northeast",1,0)
mydata$region_northwest <- ifelse(mydata$region=="northwest",1,0)
mydata$region_southeast <- ifelse(mydata$region=="southeast",1,0)
mydata <- mydata[,-6]

mydata <- as.matrix(mydata)
mydata <- as.numeric(mydata)

mydata <- matrix(mydata, ncol=9)

X <- mydata[, c(1:4,6:9)]
Y <- mydata[, 5]
```

### Init

This function gives the starting value for the predictors of the logistic regression. The specific calculation uses the least squares formula given by $(X^TX)^{−1}X^Ty$.

```{r}
(init_beta_demo <- init(resp = Y, pred= X))
```

## Boot_CI

This function will conduct a bootstrap routine for each predictor or beta value for the logistic regression curve. The user is able to choose the amount of bootstraps to conduct as well as the confidence level for the intervals. Here, we will use it on the loaded data. We will use the default values for n and alpha and X and Y for our response and predictor variables.

```{r}
(Boot_CI_demo <- Boot_CI(n=20, alpha=0.05, pred= X, resp= Y))
```

## log_plot

This function will plot the optimized logistic curve for the given data,  where the y-axis is the binary response y while the x-axis represents a sequence of values from the range of fitted values from the logistic regression.

```{r}
log_plot(pred= X, resp= Y)
```

## conf_mat

This function will create a "confusion matrix" using a cut-off value for prediction at 0.5. It will also print the Prevalence, Accuracy, Sensitivity, Specificity, False Discovery Rate, and Diagnostic Odds Ratio metrics when run which can be calculated by values within the matrix. The matrix is 3x3 with values representing the performance of the logistic regression curve that we obtain.

```{r}
conf_mat(pred = X, resp = Y)
```

## plot_metric

This function conducts logistic regression on the given data and plots a given metric that visualizes the performance of the regression over a grid of cut-off values for prediction going from 0.1 to 0.9 with steps of 0.1. The metrics that can be entered are the same as the metrics printed in conf_mat.

```{r}
plot_metric(resp=Y, pred=X, metric = "Accuracy")
```



